Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-31 curvature.tex
Coarse, not ric.
Mostly typography.

\begin{figure}
	\hspace*{\stretch{1}}
	\includegraphics[scale=1.25]{figs/two-spr}
	\hspace*{\stretch{1}}
	\caption{Two trees differing by two rSPR operations that move the grey subtrees.}
	\label{fig:two-spr}
\end{figure}

@@

Lin et al.~\cite{lin2011ricci} defined a variant definition of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
The lazy random walk selects a neighbor according to a given distribution $m$ as before, but only travels to the neighbor with probability $p$.
Thus the lazy mass assignment $m^p_x$ assigns $p\, m_x$ mass to each neighbor of $x$ and the remaining $1 - p$ mass to $x$.
We denote the coarse curvature of the $p$-lazy random walk between two vertices $x$ and $y$ with respect to a random walk $m$ by $\curvature[p]{m; x,y}$.
For example, $\curvature[\frac{1}{4}]{m; x,y}$ describes the curvature of the lazy random walk that follows the given random walk $m$ with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
The asymptotic Ricci-Ollivier curvature of $x$ and $y$ is then:
\begin{equation}
\ric{m; x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{m; x,y}}{p}
\label{eq:curvatureDef}
\end{equation}
As above for $\kappa$, we use $\ric{x, y}$ as shorthand for $\ric{m; x, y}$ when $m$ is the uniform lazy random walk, and $\ric{\MH; x, y}$ when $m$ is the Metropolis-Hastings random walk sampling the uniform distribution (Section~\ref{sec:random_walks}).
This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g. Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.

@@

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter5}}
	\hspace*{\stretch{2}}
	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
	\hspace*{\stretch{1}}
    \caption{Scatterplot of $\ric{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
	\label{fig:rspr-scatter}
\end{figure}

@@

\begin{lemma}
\label{lem:asymptotic}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
		\item	$\ric{MH; T,S} = \curvature{MH; T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{MH; T,S} \le \ric{MH; T,S} \le \curvature{MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lemma}

\begin{proof}
	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\curvature[p]{T,S} /p &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&\ge \curvature{T,S}
	\end{align*}
The lower bound in the MH case follows analogously.

>>>>

\begin{figure}
	\hspace*{\stretch{1}}
	\includegraphics[scale=1.25]{figs/two-spr}
	\hspace*{\stretch{1}}
	\caption{\
    Two rSPR operations, each of which moves one grey subtree.
    The leftmost and rightmost trees are rSPR distance two apart.}
	\label{fig:two-spr}
\end{figure}

@@

Lin et al.~\cite{lin2011ricci} defined a variant definition of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
The lazy random walk selects a neighbor according to a given distribution $m$ as before, but only travels to the neighbor with probability $p$.
Thus the lazy mass assignment $m^p_x$ assigns $p\, m_x$ mass to each neighbor of $x$ and the remaining $1 - p$ mass to $x$.
We denote the coarse curvature of the $p$-lazy random walk between two vertices $x$ and $y$ with respect to a random walk $m$ by $\curvature[p]{m; x,y}$.
For example, $\curvature[\frac{1}{4}]{m; x,y}$ describes the curvature of the lazy random walk that follows the given random walk $m$ with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
The asymptotic Ricci-Ollivier curvature of $x$ and $y$ is then:
\begin{equation}
\ric{m; x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{m; x,y}}{p}
\label{eq:asympCurvatureDef}
\end{equation}
As above for $\kappa$, we use $\ric{x, y}$ as shorthand for $\ric{m; x, y}$ when $m$ is the uniform lazy random walk, and $\ric{\MH; x, y}$ when $m$ is the Metropolis-Hastings random walk sampling the uniform distribution (Section~\ref{sec:random_walks}).
This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g. Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.

@@

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter5}}
	\hspace*{\stretch{2}}
	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
	\hspace*{\stretch{1}}
    \caption{Scatterplot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Color displays the average degree of $T_1$ and $T_2$. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
	\label{fig:rspr-scatter}
\end{figure}

@@

\begin{lemma}
\label{lem:asymptotic}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
		\item	$\ric{\MH; T,S} = \curvature{\MH; T,S}$, if $\dspr{T,S} > 1$, and
		\item	$\curvature{\MH; T,S} \le \ric{\MH; T,S} \le \curvature{\MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lemma}

\begin{proof}
	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\curvature[p]{T,S} /p &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
		&= \left(\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}\right) \\
		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&= \curvature{T,S}
	\end{align*}
The lower bound in the MH case follows analogously.
